---
title: "HendricksonSwen_A8"
author: "Swen Hendrickson"
date: "11/15/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---
# Difference-in-Differences Analysis on BART Ridership 
   (Assessing Nearby Towns to New (2018) Antioch BART Station)






---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, message = F, warning = F)

# Clear all variables, close all plots, and clear the console, load libraries
rm(list=ls())   # clears variables
graphics.off()   # clears plots
cat("\014")   # clears console

setwd("~/GitHub/swehen17.github.io/218X/A8")

library(tigris)
library(tidyverse)
library(tidycensus)
library(sf)
library(censusapi)
library(leaflet)
library(StatMatch)
```

```{r SETUP PUMAS}
Sys.setenv(CENSUS_KEY="c8aa67e4086b4b5ce3a8717f59faa9a28f611dab")

ca_pumas <-
  pumas("CA", cb = T, progress_bar = F)

bay_county_names <-
  c(
    "Alameda",
    "Contra Costa",
    "Marin",
    "Napa",
    "San Francisco",
    "San Mateo",
    "Santa Clara",
    "Solano",
    "Sonoma"
  )

bay_counties <-
  counties("CA", cb = T, progress_bar = F) %>%
  filter(NAME %in% bay_county_names)

bay_pumas <-
  ca_pumas %>% 
  st_centroid() %>% 
  .[bay_counties, ] %>% 
  st_drop_geometry() %>% 
  left_join(ca_pumas %>% select(GEOID10)) %>% 
  st_as_sf()
```

Location of new Antioch station and nearby PUMA of interest (which comprises towns of Oakley and Brentwood):

```{r MAP IT WITH BART STATION AND PUMA OF INTEREST}
leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    data = bay_pumas,
    weight = 1,
    color = "gray",
    label = ~PUMACE10
  ) %>% 
  addMarkers(
    lng = -121.78025199689912,   # finding lat long on google maps
    lat = 37.99593068104423       # that's the BART station
  ) %>% 
  addPolygons(
    data = bay_pumas %>% 
      filter(PUMACE10 == "01309")
  )
```

```{r BUILD DATA TO COMPARE YEARS}
# this line shortens a much longer to load chunk from textbook
pums_2014_2019 <- readRDS("pums_2014_2019_wts.rds")

pums_bart <- pums_2014_2019 %>%
  mutate(
    PWGTP = as.numeric(PWGTP),
    bart = ifelse(
      JWTR %in% c("4"),
      PWGTP,
      0
    )
  ) %>% 
  group_by(PUMA, year) %>% 
  summarize(
    pop = sum(PWGTP),
    bart = sum(bart)
  )
```

2017 Bay Area PUMA Populations (for reference):

```{r 2017 VISUALIZATION OF POPULATION for reference}
# Population first
pums_pal <- colorNumeric(
  palette = "YlOrRd",
  domain = pums_bart %>% 
    filter(year == 2017) %>% 
    pull(pop)
)

leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    data = pums_bart %>% 
      filter(year == 2017) %>% 
      right_join(bay_pumas %>% select(PUMA = PUMACE10)) %>% 
      st_as_sf(),
    fillColor = ~pums_pal(pop),
    color = "white",
    weight = 1,
    fillOpacity = 0.5,
    label = ~paste0(PUMA,": Population ", pop)
  )
```

2017 Bay Area BART Ridership (for reference):

```{r 2017 VISUALIZATION OF BART RIDERSHIP for reference}
#
# Now BART Ridership
#
pums_pal <- colorNumeric(
  palette = "GnBu",
  domain = pums_bart %>% 
    filter(year == 2017) %>% 
    pull(bart)
)

leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    data = pums_bart %>% 
      filter(year == 2017) %>% 
      right_join(bay_pumas %>% select(PUMA = PUMACE10)) %>% 
      st_as_sf(),
    fillColor = ~pums_pal(bart),
    color = "white",
    weight = 1,
    fillOpacity = 0.5,
    label = ~paste0(PUMA,": ", bart, " BART commute riders")
  )
```

Our focus PUMA (Oakley/Brentwood) in Red with 10 most similar PUMAs (in terms of BART ridership before 2018) in Blue:

```{r DATA MANIPULATIONS and MATCHING}
# pivot to wide format
pums_bart_clean <-
  pums_bart %>% 
  select(-pop) %>% 
  pivot_wider(       # pivot wider key function here, for easier math comparison
    names_from = year,
    values_from = bart
  )

# stuff from Ch 6.1
obs_matrix <-
  pums_bart_clean %>% 
  ungroup() %>% 
  select(`2014`,`2015`,`2016`, `2017`) %>%   # add 2017
    as.matrix()   

dist_matrix <- mahalanobis.dist(obs_matrix)

rownames(dist_matrix) <- pums_bart_clean$PUMA
colnames(dist_matrix) <- pums_bart_clean$PUMA

# find 10 most similar PUMAs to treatment PUMA (Ch 6.1 techniques)
# matching according to *absolute* BART ridership values
match <- dist_matrix["01309",] %>% 
  as.data.frame() %>% 
  rownames_to_column() %>% 
  rename(
    PUMA = rowname,
    match = "."
  ) %>% 
  right_join(
    pums_bart_clean
  ) %>% 
  arrange(match) %>% 
  .[1:11, ] %>% 
  left_join(bay_pumas %>% select(PUMA = PUMACE10)) %>% 
  st_as_sf()

# top 10 most similar pumas to given one (Oakley/Brentwood, adjacent to Antioch), based on four years
# map above
leaflet() %>% 
  addTiles() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    data = match[1, ],
    color = "red",
    label = ~PUMA
  ) %>% 
  addPolygons(
    data = match[-1, ],
    label = ~PUMA
  )
```

BART Ridership for PUMA of interest and *AVERAGE of 10 simlar PUMAs from 2014 to 2019 (new Antioch station opened in 2018 marked by dashed line):

```{r COMPARE OVER TIME}
# seeing how similar..
match_pumas <-
  match %>% 
  filter(!PUMA %in% c("01309")) %>% 
  st_drop_geometry() %>% 
  select(-match) %>% 
  pivot_longer(      # pivot longer makes 'tidier'
    -PUMA,
    names_to = "year",
    values_to = "bart"
  ) %>%
  group_by(
    year
  ) %>% 
  summarize(
    bart = mean(bart),
    PUMA = "Similar PUMAs"
  )

treatment_pumas <-    # just taking first row of data
  match %>% 
  filter(PUMA %in% c("01309")) %>% 
  select(-match) %>% 
  st_drop_geometry() %>% 
  pivot_longer(
    -PUMA,
    names_to = "year",
    values_to = "bart"
  )

rbind(      # by fields, different orders don't matter
  treatment_pumas,
  match_pumas
) %>% 
  ggplot(
    aes(
      x = as.numeric(year),   # don't want as factor here
      y = bart,
      color = PUMA
    )
  ) +
  geom_line() +
  geom_vline(xintercept = 2018, linetype = "dashed") + # to put vertical line, just check units
  labs(
    title = "Oakley/Brentwood vs. control neighborhoods, BART ridership",
    x = "Year",
    y = "BART commute riders"
  )
```

   It's important to remember here that the blue line above is an average of BART
ridership values for 10 different counties that had similar (i.e. quite low)
ridership to Oakley/Brentwood before the new 2018 Antioch station, and that 
averaging likely flattens out more bumpy individual variations. With the red line,
we see as expected (and presumably as hoped by urban planners and BART itself), a 
significant jump in ridership for the adjacent PUMA to the new station. 
   
   Some key differences between this analysis and the one given in the textbook:
   
    - Here we had one extra year of 'pre new station' data and one less year of
      'post new station' data. This perhaps added some strength in the matching
      between our PUMA and its 'comparables' (to use as proxies for experimental            controls). Also, given only one year of post new station data, we don't have
      much time in our data to see if the bump in ridership evened out, or better put,
      whether ridership in the 'similar' PUMAs caught up with ridership nearby 
      where the new station was put in.

    - Also, it should be noted that the scale of the Y-axis in this analysis spans
      about twice the ridership of that seen in the textbook graph. In fact the
      change seen in the blue line (ridership for the 10 control PUMAs) is similar
      in both analyses (and in fact going back to visually inspect the Leaflet
      maps in each case reveals that most of these control PUMAs are the same 
      in both cases with just a couple that differ). These PUMAs seem to be 
      the Bay Area PUMAs with the absolute least amount of BART ridership, as two
      PUMAs actually have 0 ridership for all 6 years considered, and a few other
      PUMAs have 0 for most years here. Considering the focus PUMA now (Oakley/
      Brentwood), our new graph shows a roughly similar proportional spike in
      ridership (about 3x), but starting from a significantly higher amount (500
      instead of 50). Interestingly, ridership also drops slightly in the year
      immediately following the new station opening.

    - It's important to reiterate some of the disclaimers explained in the text
      book, like the face that larger datasets could be sought out in order to 
      reduce noice (and increase statistical significance), that additional 
      variables could have been used to match with our control PUMAs, that we 
      could have chosen more or fewer than the arbitrary number of 10, that the 
      geographical boundary chosen (our PUMA) might have better alternatives
      (potentially smaller potentially larger), that our selection of years is
      limited by PUMS data available, that we made assumptions about people 
      answering the ACS survey in a certain way (to lead to reliable numbers
      for BART ridership), and that other impacts of a new BART station could
      have been evaluated and could be more significant in terms of social
      impact and urban systems.
   
DiD Regression Summary:
   
```{r DiD REGRESSION}
# then make Diff in Diff regression
# time is just 0's and 1's
transit_did <-
  match %>% 
  st_drop_geometry() %>% 
  select(-match) %>% 
  pivot_longer(
    -PUMA,
    names_to = "year",
    values_to = "bart"
  ) %>% 
  mutate(
    year = year %>% as.numeric(),
    time = ifelse(year >= 2018, 1, 0),
    treated = ifelse(PUMA == "01309", 1, 0)
  )

did_reg <- lm(bart ~ treated * time, data = transit_did)

summary(did_reg)
```

Our DiD regression here shows a substantial effect on ridership of the new
Antioch station with an increase of 1097 (riders) as compared to the controls.
This figure is read from the 'treated:time' coefficient listed last in the
summary (denoting the rate of change of the 'treated' group when controlling
for the effect of time). Importantly, this coefficient is listed with very high 
statistical significance (three ***'s for 99% confidence).























