---
title: "HendricksonSwen_Project"
author: "Swen Hendrickson"
date: "12/1/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r 1. setup, include = F}
knitr::opts_chunk$set(echo = F, warning = F, message = F)

# Clear all variables, close all plots, and clear the console, load libraries
rm(list=ls())
graphics.off()
cat("\014")

setwd("~/GitHub/swehen17.github.io/218X/Project")

library(censusapi)
library(tidyverse)
library(tigris)
library(sf)
library(leaflet)
library(mapview)
library(plotly)
library(data.table)
library(viridis)
library(hrbrthemes)

Sys.setenv(CENSUS_KEY="c8aa67e4086b4b5ce3a8717f59faa9a28f611dab")

acs_vars_2019_5yr <- readRDS("acs_vars_2019_5yr.rds")
```

(methodology below)
```{r 2.1 DATA-IN: ENERGY INTENSITY per SQUARE FOOT [EISF]}
residential_energy_intensity <- read_csv("usage_bld_btu.csv") %>%
  filter(
    usetype == "single_family",
  ) %>%
  select(
    geo_id,
    usetype,
    year,
    sqft,
    usage,
    name
  )

# only include census tracts, remove other geographical aggregations (cities, zips, etc.)
residential_energy_intensity <- residential_energy_intensity[1:7920,]

# sum(is.na(residential_energy_intensity$usage))
# 459 NA values for usage

# remove 'masked' values and create energy intensity calc
residential_energy_intensity <- residential_energy_intensity %>%
  filter(
    usage != "masked"
  ) %>%
  mutate(
    energy_intensity = as.numeric(usage) / sqft
  )

# map energy intensity
#

ca_tracts <- readRDS("ca_tracts.rds") %>%
  st_transform(crs= 4326)

# next steps: follow steps in A5

res_eng_int_2017 <- residential_energy_intensity %>%
  filter(year == 2017)

# get just tracts in my energy dataset
bay_tracts_need <- unique(residential_energy_intensity$name)

saveRDS(bay_tracts_need, file = "bay_tracts_need.")

bay_tracts_use_geo <- ca_tracts %>%
  filter(GEOID %in% bay_tracts_need) %>%
  select(GEOID, geometry) %>%
  transmute(name = GEOID) %>%
  st_transform(crs = 4326)

bay_eng_int_map <- residential_energy_intensity %>%
  left_join(bay_tracts_use_geo, by = "name") %>%
  select(year, energy_intensity, geometry, name) %>%
  st_as_sf() %>%
  filter(energy_intensity != "Inf") %>%
  st_transform(crs = 4326)

saveRDS(bay_eng_int_map, "bay_eng_int_map.rds")

beim_2013 <- bay_eng_int_map %>%
  filter(year == 2013) %>%
  st_transform(crs = 4326)

saveRDS(beim_2013, "beim_2013.rds")

# actually mapping, finally
# 2013
beim_pal <- colorNumeric(
  palette = "Oranges",
  domain = bay_eng_int_map$energy_intensity
)

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    data = beim_2013,
    fillColor = ~beim_pal(energy_intensity),
    color = "white",
    weight = 0.5,
    fillOpacity = 0.5,
    label = ~energy_intensity
  )

# avg of 2013-2017
beim_2013to2017_avgs <- as.data.frame(bay_eng_int_map) %>%
  group_by(
    name
  ) %>%
  summarise(
    avg = sum(energy_intensity / 5)
  ) %>% left_join(
    bay_tracts_use_geo, by = "name"
  ) %>%
  st_as_sf()

# mapping avgs now
# 2013-2017
beim_pal <- colorNumeric(
  palette = "Blues",
  domain = beim_2013to2017_avgs$avg
)

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    data = beim_2013to2017_avgs,
    fillColor = ~beim_pal(avg),
    color = "white",
    weight = 0.5,
    fillOpacity = 0.5,
    label = ~avg
  )
```
Methodology of above chunk:
	- Took in data (from Energy Atlas) which included utility energy consumption 
	data for buildings of all types, aggregated by census tract
	- Filtered for residential single-family homes (data included multi family, 
	commercial, industrial, etc.)
	- Divided total usage by built SF to get energy intensity
	- mapped an individual year, then
	- Took average of year 2013-2017
  - Mapped by energy intensity (Btu/SF), darker blue being higher

```{r 2.2 DATA-IN: ENERGY CONSUMPTION PER CAPITA [ECPC]}
# now get energy consumption per capita (from the same data set [Energy Atlas])
ECPC <- read_csv("usage_bld_btu.csv") %>%
  select(usetype, year, sqft, usage_percap, name) %>%
  filter(
    usetype %in% c('res_total', 'single_family'),
    name %in% unique(residential_energy_intensity$name)) %>%
  mutate(
    singfam_perc = sqft / shift(sqft, 1L, type = "lag"),
    usage_percap = singfam_perc * shift(usage_percap, 1L, type = "lag")
    ) %>%
  filter(usetype == "single_family") %>%
  select(year, usage_percap, name)

# TODO: map this like above (each year, and then average)
```
Methodology of above:
- only had population for total residential (includes mutli family, etc)
- but had SF data for single family homes. got percentage that single family
homes make up of total residential, to multiply by the usage per capita for
total residential to get usage per capita for single family homes
- helps to control for neighborhoods that might have different proportions 
of homes vs. multifamily complexes, mobile homes, etc.

```{r 3.1 DATA-IN: INCOME}
# ACS 5-year estimates for years 2013 - 2017
income_2013 <-
  read_csv("data/income/ACSDT5Y2013.B19001_data_with_overlays_2021-12-04T162248.csv") %>%
  mutate(name = substr(GEO_ID, 10, 20)) %>%
  select(-c(ends_with("M"), "NAME", "GEO_ID")) %>%
  filter(name %in% unique(residential_energy_intensity$name)) %>%
  transmute(
    'total' = as.numeric(B19001_001E),
    '5k' = as.numeric(B19001_002E),
    '12.5k' = as.numeric(B19001_003E),
    '17.5k' = as.numeric(B19001_004E),
    '22.5k' = as.numeric(B19001_005E),
    '27.5k' = as.numeric(B19001_006E),
    '32.5k' = as.numeric(B19001_007E),
    '37.5k' = as.numeric(B19001_008E),
    '42.5k' = as.numeric(B19001_009E),
    '47.5k' = as.numeric(B19001_010E),
    '55k' = as.numeric(B19001_011E), # PATTERN STARTS CHANGING
    '67.5k' = as.numeric(B19001_012E),
    '87.5k' = as.numeric(B19001_013E),
    '112.5k' = as.numeric(B19001_014E),
    '137.5k' = as.numeric(B19001_015E),
    '175k' = as.numeric(B19001_016E),
    '250k' = as.numeric(B19001_017E), # $200k 'or 'more'. is there a better est for this?
    name = name) %>%
  mutate(
    weighted_avg_income =
      1000 * (`5k` * 5 +
    `12.5k` * 12.5 +
    `17.5k` * 17.5 +
    `22.5k` * 22.5 +
    `27.5k` * 27.5 +
    `32.5k` * 32.5 +
    `37.5k` * 37.5 +
    `42.5k` * 42.5 +
    `47.5k` * 47.5 +
    `55k` * 55 +
    `67.5k` * 67.5 +
    `87.5k` * 87.5 +
    `112.5k` * 112.5 +
    `137.5k` * 137.5 +
    `175k` * 175 +
    `250k` * 250)
    / total)

##

income_2014 <-
  read_csv("data/income/ACSDT5Y2014.B19001_data_with_overlays_2021-12-04T162248.csv") %>%
  mutate(name = substr(GEO_ID, 10, 20)) %>%
  select(-c(ends_with("M"), "NAME", "GEO_ID", "B19001_001E")) %>%
  filter(name %in% unique(residential_energy_intensity$name))

income_2015 <-
  read_csv("data/income/ACSDT5Y2015.B19001_data_with_overlays_2021-12-04T162248.csv") %>%
  mutate(name = substr(GEO_ID, 10, 20)) %>%
  select(-c(ends_with("M"), "NAME", "GEO_ID", "B19001_001E")) %>%
  filter(name %in% unique(residential_energy_intensity$name))

income_2016 <-
  read_csv("data/income/ACSDT5Y2016.B19001_data_with_overlays_2021-12-04T162248.csv") %>%
  mutate(name = substr(GEO_ID, 10, 20)) %>%
  select(-c(ends_with("M"), "NAME", "GEO_ID", "B19001_001E")) %>%
  filter(name %in% unique(residential_energy_intensity$name))

income_2017 <-
  read_csv("data/income/ACSDT5Y2017.B19001_data_with_overlays_2021-12-04T162248.csv") %>%
  mutate(name = substr(GEO_ID, 10, 20)) %>%
  select(-c(ends_with("M"), "NAME", "GEO_ID", "B19001_001E")) %>%
  filter(name %in% unique(residential_energy_intensity$name))
```

```{r 3.2 DATA-IN: OTHER}
# Houshold Size

# Tenure (Own vs. Rent)

# Number of Years Living in Home

# Other?
```

```{r 3.3 DATA-IN: RACE}
race_2010 <-
  read_csv(
    "data/race/DECENNIALSF12010.P8_data_with_overlays_2021-12-05T025058.csv")
race_2010 <- race_2010[2:8058 , 1:10] %>%
  mutate(name = substr(GEO_ID, 10, 20)) %>% select(-c(GEO_ID, P008002)) %>%
  transmute(
    'Total' = as.numeric(P008001),
    'White' = as.numeric(P008003),
    'Black' = as.numeric(P008004),
    'Native_American' = as.numeric(P008005),
    'Asian' = as.numeric(P008006),
    'Pacific_Islander' = as.numeric(P008007),
    'Other' = as.numeric(P008008),
    'Multi-Racial' = as.numeric(P008009),
    'name' = name
  ) %>%
  filter(name %in% unique(residential_energy_intensity$name))

race_2010 <- race_2010 %>%
  transmute(
    'White' = White / Total,
    'Black' = Black / Total,
    'Native_American' = Native_American / Total,
    'Asian' = Asian / Total,
    'Pacific_Islander' = Pacific_Islander / Total,
    'Other' = Other / Total,
    'Multi-Racial' = `Multi-Racial` / Total,
    'name' = name
  ) %>% filter(!is.na(.))
# this gives us census tracts with race breakdowns by percentage
```

```{r 4.1 ANALYSIS: EISF vs. INCOME}
# plot energy intensity by income
# 2013
income_engint_2013 <- beim_2013 %>%
  left_join(income_2013, by = "name") %>%
  select(weighted_avg_income, energy_intensity) %>%
  st_drop_geometry()

ggplot(
    data = income_engint_2013,
    aes(
      x = weighted_avg_income,
      y = energy_intensity
    )
  ) +
  geom_point() +
  ggtitle("Energy Intensity vs. Income") +
  xlab("Weighted Average Household Income (USD)") +
  ylab("Energy Intensity (Btu/SF)") +
  geom_smooth(method = "lm")

# color code low EISF cluster
cluster <- income_engint_2013 %>%
  filter(energy_intensity < 25000)

ggplot(
    data = income_engint_2013,
    aes(
      x = weighted_avg_income,
      y = energy_intensity
    )
  ) +
  geom_point() +
  ggtitle("Energy Intensity vs. Income") +
  xlab("Weighted Average Household Income (USD)") +
  ylab("Energy Intensity (Btu/SF)") +
  # geom_smooth(method = "lm") +
  geom_point(data = cluster, color = "purple")

# TO UPDATE TITLE FORMATTING:
# theme_update(plot.title = element_text(size = 22))
# theme(plot.title = element_text(hjust = 0.5)) +
#

# as income goes up, energy intensity per SF goes down
# cluster of low intensity bldgs
# TODO: look into where these buildings are

model <- lm(energy_intensity ~ weighted_avg_income, data = income_engint_2013)

summary(model)

# where are the super efficient tracts?
efficient_tracts_2013 <- beim_2013 %>%
  filter(energy_intensity < 25000)
  ###uneeq

# map it
beim_pal <- colorNumeric(
  palette = "Green",
  domain = efficient_tracts_2013$energy_intensity
)

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    data = efficient_tracts_2013,
    fillColor = ~beim_pal(energy_intensity),
    color = "white",
    weight = 0.5,
    fillOpacity = 0.5,
    label = ~energy_intensity
  )
# noticeably, big urban centers are missing (SF, Oak, San Jose)
```

```{r 4.2 ANALYSIS: ECPC vs. INCOME}
# plot ECPC by income
# 2013
income_ECPC_2013 <- ECPC %>%
  filter(year == 2013, !is.na(usage_percap)) %>%
  left_join(income_2013, by = "name") %>%
  select(weighted_avg_income, usage_percap)

ggplot(
    data = income_ECPC_2013,
    aes(
      x = weighted_avg_income,
      y = usage_percap
    )
  ) +
  geom_point() +
  ggtitle("Energy Consumption Per Capita vs. Income") +
  xlab("Weighted Average Household Income (USD)") +
  ylab("Energy Consumption Per Capita (Btu/Person)") +
  geom_smooth(method = "lm")

# as income goes up, consumption per capita increases steadily!

model <- lm(usage_percap ~ weighted_avg_income, data = income_ECPC_2013)

summary(model)
```

```{r 5.1 ANALYSIS: EISF vs. OTHER VARS}
# replace with analysis of new vars (not education)
# (see chunk 5: 3.2)
```

```{r 5.2 ANALYSIS: ECPC vs. EDUCATION}
# merp
```




# Equity Analysis
   Equity analysis in this case is tricky, since consumption data from the utility (PG&E) was, for privacy reasons, aggregated to the census tract level as the smallest geographical granularity. This means that we can compare a few census tracts by their overall efficiency scores (EISF for single family homes) and their overall racial makeup, but we cannot say for certain which specific houses operate more or less efficiently and who exactly lives in those houses.
  So, the tracts with both the highest and the lowest EISF scores (shown in the EISF maps) will be inspected for their racial make up to see if any trends are apparent.
   
```{r 6.0: ANALYSIS: EQUITY ANALYSIS (RACE)}
# find best and worst tracts by EISF (using 2013-2017 avg)
 # data: beim_2013to2013_avgs
# then join with race_2010 by name to produce pie charts

# MOST energy efficient tract:
lowest_EISF_tract_2013to2017 <- beim_2013to2017_avgs %>% 
  filter(name == "06081613501") %>% 
  left_join(race_2010, by = "name")
# map
mapview(lowest_EISF_tract_2013to2017)
# join w/ race data
data1 <- lowest_EISF_tract_2013to2017[ , 4:10] %>% st_drop_geometry()
# plot
data <- data.frame(
  name = c(colnames(data1)) ,  
  value = c(as.numeric(data1[1, ]))
  )
# ..
ggplot(data, aes(x = name, y = value, fill = name)) + 
  geom_bar(stat = "identity") +
  ggtitle("Tract with LOWEST EISF") +
  xlab("Race Category") +
  ylab("Percentage of Population")


# LEST efficient tract:
highest_EISF_tract_2013to2017 <- beim_2013to2017_avgs %>% 
  filter(name == "06013314104") %>% 
  left_join(race_2010, by = "name")
# map
mapview(highest_EISF_tract_2013to2017)
# join w/ race data
data1 <- highest_EISF_tract_2013to2017[ , 4:10] %>% st_drop_geometry()
# plot
data <- data.frame(
  name = c(colnames(data1)) ,  
  value = c(as.numeric(data1[1, ]))
  )
# ..
ggplot(data, aes(x = name, y = value, fill = name)) + 
  geom_bar(stat = "identity") +
  ggtitle("Tract with HIGHEST EISF") +
  xlab("Race Category") +
  ylab("Percentage of Population")

# TODO: could also find Q1 and Q3 (circle back later)
```

  As we can see above..


---

Scrap Code Below:

```{r scrap code 1}
# bay_education_income_tract <- 
#   getCensus(
#     name = "acs/acs5",
#     vintage = 2019,
#     region = "tract:*",
#     regionin = "state:06+county:001,013,041,055,075,081,085,095,097",
#     vars = c(
#       "B06009_001E",   # educational attainment
#       "B06009_002E",
#       "B06009_003E",
#       "B19001_001E",   # income
#       "B19001_014E",
#       "B19001_015E",
#       "B19001_016E",
#       "B19001_017E"
#     )
#   ) %>% 
#   transmute(
#     tract = paste0(state, county, tract),
#     perc_college = 1 - (B06009_002E + B06009_003E) / B06009_001E,
#     perc_over100k = (B19001_014E + B19001_015E + B19001_016E + B19001_017E) / B19001_001E
#   ) %>% 
#   filter(
#     !is.na(perc_college), 
#     !is.na(perc_over100k)
#   )
```

```{r scrap code 2}
# census_race_categories <- 
#   c(
#     "White Alone",
#     "Black or African American",
#     "American Indian and Alaska Native Alone",
#     "Asian Alone",
#     "Native Hawaiian and Other Pacific Islander Alone",
#     "Some Other Race Alone",
#     "Two or More Races"
#   )
# 
# bay_income_race <-
#   1:7 %>% 
#   map_dfr(function(x){
#     getCensus(
#       name = "acs/acs5",
#       vintage = 2019,
#       region = "county:001,013,041,055,075,081,085,095,097",
#       regionin = "state:06",
#       vars = paste0("group(B19001",LETTERS[x],")")
#     ) %>%
#       select(!c(GEO_ID,state,NAME) & !ends_with(c("EA","MA","M"))) %>%
#       pivot_longer(
#         ends_with("E"),
#         names_to = "name",
#         values_to = "estimate"
#       ) %>%
#       left_join(
#         acs_vars_2019_5yr %>% 
#           select(name, label)
#       ) %>% 
#       select(-name) %>% 
#       separate(
#         label,
#         into = c(NA,NA,"income"),
#         sep = "!!"
#       ) %>% 
#       filter(!is.na(income)) %>% 
#       mutate(race = census_race_categories[x])
#   })
```



















