---
title: "HendricksonSwen_Project_DATA-IN"
author: "Swen Hendrickson"
date: "12/12/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r 1. setup, include = F}
knitr::opts_chunk$set(echo = F, warning = F, message = F)

# Clear all variables, close all plots, and clear the console, load libraries
rm(list=ls())
graphics.off()
cat("\014")

setwd("~/GitHub/swehen17.github.io/218X/Project")

library(censusapi)
library(tidyverse)
library(tigris)
library(sf)
library(leaflet)
library(mapview)
library(plotly)
library(data.table)
library(viridis)
library(hrbrthemes)
library(janitor)

Sys.setenv(CENSUS_KEY="c8aa67e4086b4b5ce3a8717f59faa9a28f611dab")
```

Pasted in from v2:
(methodology below)
```{r 2.1 DATA-IN: ENERGY INTENSITY per SQUARE FOOT [EISF]}
residential_energy_intensity <- read_csv("usage_bld_btu.csv") %>%
  filter(
    usetype == "single_family",
  ) %>%
  select(
    geo_id,
    usetype,
    year,
    sqft,
    usage,
    name
  )

# only include census tracts, remove other geographical aggregations (cities, zips, etc.)
residential_energy_intensity <- residential_energy_intensity[1:7920,]

# sum(is.na(residential_energy_intensity$usage))
# 459 NA values for usage

# remove 'masked' values and create energy intensity calc
residential_energy_intensity <- residential_energy_intensity %>%
  filter(
    usage != "masked"
  ) %>%
  mutate(
    energy_intensity = as.numeric(usage) / sqft
  )

# map energy intensity
#

ca_tracts <- readRDS("ca_tracts.rds") %>%
  st_transform(crs= 4326)

# next steps: follow steps in A5

res_eng_int_2017 <- residential_energy_intensity %>%
  filter(year == 2017)

# get just tracts in my energy dataset
bay_tracts_need <- unique(residential_energy_intensity$name)

saveRDS(bay_tracts_need, file = "bay_tracts_need.rds")

bay_tracts_use_geo <- ca_tracts %>%
  filter(GEOID %in% bay_tracts_need) %>%
  select(GEOID, geometry) %>%
  transmute(name = GEOID) %>%
  st_transform(crs = 4326)

bay_eng_int_map <- residential_energy_intensity %>%
  left_join(bay_tracts_use_geo, by = "name") %>%
  select(year, energy_intensity, geometry, name) %>%
  st_as_sf() %>%
  filter(energy_intensity != "Inf") %>%
  st_transform(crs = 4326)

saveRDS(bay_eng_int_map, "bay_eng_int_map.rds")

beim_2013 <- bay_eng_int_map %>%
  filter(year == 2013) %>%
  st_transform(crs = 4326)

saveRDS(beim_2013, "beim_2013.rds")

# actually mapping, finally
# 2013
beim_pal <- colorNumeric(
  palette = "Oranges",
  domain = bay_eng_int_map$energy_intensity
)

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    data = beim_2013,
    fillColor = ~beim_pal(energy_intensity),
    color = "white",
    weight = 0.5,
    fillOpacity = 0.5,
    label = ~energy_intensity
  )

# avg of 2013-2017
beim_2013to2017_avgs <- as.data.frame(bay_eng_int_map) %>%
  group_by(
    name
  ) %>%
  summarise(
    avg = sum(energy_intensity / 5)
  ) %>% left_join(
    bay_tracts_use_geo, by = "name"
  ) %>%
  st_as_sf()

# mapping avgs now
# 2013-2017
beim_pal <- colorNumeric(
  palette = "Blues",
  domain = beim_2013to2017_avgs$avg
)

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    data = beim_2013to2017_avgs,
    fillColor = ~beim_pal(avg),
    color = "white",
    weight = 0.5,
    fillOpacity = 0.5,
    label = ~avg
  )
```
Methodology of above chunk:
	- Took in data (from Energy Atlas) which included utility energy consumption 
	data for buildings of all types, aggregated by census tract
	- Filtered for residential single-family homes (data included multi family, 
	commercial, industrial, etc.)
	- Divided total usage by built SF to get energy intensity
	- mapped an individual year, then
	- Took average of year 2013-2017
  - Mapped by energy intensity (Btu/SF), darker blue being higher

```{r 2.2 DATA-IN: ENERGY CONSUMPTION PER CAPITA [ECPC]}
# now get energy consumption per capita (from the same data set [Energy Atlas])
ECPC <- read_csv("usage_bld_btu.csv") %>%
  select(usetype, year, sqft, usage_percap, name) %>%
  filter(
    usetype %in% c('res_total', 'single_family'),
    name %in% unique(residential_energy_intensity$name)) %>%
  mutate(
    singfam_perc = sqft / shift(sqft, 1L, type = "lag"),
    usage_percap = singfam_perc * shift(usage_percap, 1L, type = "lag")
    ) %>%
  filter(usetype == "single_family") %>%
  select(year, usage_percap, name)

# TODO: map this like above (each year, and then average)
```
Methodology of above:
- only had population for total residential (includes mutli family, etc)
- but had SF data for single family homes. got percentage that single family
homes make up of total residential, to multiply by the usage per capita for
total residential to get usage per capita for single family homes
- helps to control for neighborhoods that might have different proportions 
of homes vs. multifamily complexes, mobile homes, etc.

```{r 3.1 DATA-IN: INCOME}
# ACS 5-year estimates for years 2013 - 2017
income_2013 <-
  read_csv("data/income/ACSDT5Y2013.B19001_data_with_overlays_2021-12-04T162248.csv") %>%
  mutate(name = substr(GEO_ID, 10, 20)) %>%
  select(-c(ends_with("M"), "NAME", "GEO_ID")) %>%
  filter(name %in% unique(residential_energy_intensity$name)) %>%
  transmute(
    'total' = as.numeric(B19001_001E),
    '5k' = as.numeric(B19001_002E),
    '12.5k' = as.numeric(B19001_003E),
    '17.5k' = as.numeric(B19001_004E),
    '22.5k' = as.numeric(B19001_005E),
    '27.5k' = as.numeric(B19001_006E),
    '32.5k' = as.numeric(B19001_007E),
    '37.5k' = as.numeric(B19001_008E),
    '42.5k' = as.numeric(B19001_009E),
    '47.5k' = as.numeric(B19001_010E),
    '55k' = as.numeric(B19001_011E), # PATTERN STARTS CHANGING
    '67.5k' = as.numeric(B19001_012E),
    '87.5k' = as.numeric(B19001_013E),
    '112.5k' = as.numeric(B19001_014E),
    '137.5k' = as.numeric(B19001_015E),
    '175k' = as.numeric(B19001_016E),
    '250k' = as.numeric(B19001_017E), # $200k 'or 'more'. is there a better est for this?
    name = name) %>%
  mutate(
    weighted_avg_income =
      1000 * (`5k` * 5 +
    `12.5k` * 12.5 +
    `17.5k` * 17.5 +
    `22.5k` * 22.5 +
    `27.5k` * 27.5 +
    `32.5k` * 32.5 +
    `37.5k` * 37.5 +
    `42.5k` * 42.5 +
    `47.5k` * 47.5 +
    `55k` * 55 +
    `67.5k` * 67.5 +
    `87.5k` * 87.5 +
    `112.5k` * 112.5 +
    `137.5k` * 137.5 +
    `175k` * 175 +
    `250k` * 250)
    / total)

##

income_2014 <-
  read_csv("data/income/ACSDT5Y2014.B19001_data_with_overlays_2021-12-04T162248.csv") %>%
  mutate(name = substr(GEO_ID, 10, 20)) %>%
  select(-c(ends_with("M"), "NAME", "GEO_ID", "B19001_001E")) %>%
  filter(name %in% unique(residential_energy_intensity$name))

income_2015 <-
  read_csv("data/income/ACSDT5Y2015.B19001_data_with_overlays_2021-12-04T162248.csv") %>%
  mutate(name = substr(GEO_ID, 10, 20)) %>%
  select(-c(ends_with("M"), "NAME", "GEO_ID", "B19001_001E")) %>%
  filter(name %in% unique(residential_energy_intensity$name))

income_2016 <-
  read_csv("data/income/ACSDT5Y2016.B19001_data_with_overlays_2021-12-04T162248.csv") %>%
  mutate(name = substr(GEO_ID, 10, 20)) %>%
  select(-c(ends_with("M"), "NAME", "GEO_ID", "B19001_001E")) %>%
  filter(name %in% unique(residential_energy_intensity$name))

income_2017 <-
  read_csv("data/income/ACSDT5Y2017.B19001_data_with_overlays_2021-12-04T162248.csv") %>%
  mutate(name = substr(GEO_ID, 10, 20)) %>%
  select(-c(ends_with("M"), "NAME", "GEO_ID", "B19001_001E")) %>%
  filter(name %in% unique(residential_energy_intensity$name))
```

```{r 3.2 DATA-IN: OTHER}
# Houshold Size

# Tenure (Own vs. Rent)

# Number of Years Living in Home

# Other?
```

```{r 3.3 DATA-IN: RACE}
race_2010 <-
  read_csv(
    "data/race/DECENNIALSF12010.P8_data_with_overlays_2021-12-05T025058.csv")
race_2010 <- race_2010[2:8058 , 1:10] %>%
  mutate(name = substr(GEO_ID, 10, 20)) %>% select(-c(GEO_ID, P008002)) %>%
  transmute(
    'Total' = as.numeric(P008001),
    'White' = as.numeric(P008003),
    'Black' = as.numeric(P008004),
    'Native_American' = as.numeric(P008005),
    'Asian' = as.numeric(P008006),
    'Pacific_Islander' = as.numeric(P008007),
    'Other' = as.numeric(P008008),
    'Multi-Racial' = as.numeric(P008009),
    'name' = name
  ) %>%
  filter(name %in% unique(residential_energy_intensity$name))

race_2010 <- race_2010 %>%
  transmute(
    'White' = White / Total,
    'Black' = Black / Total,
    'Native_American' = Native_American / Total,
    'Asian' = Asian / Total,
    'Pacific_Islander' = Pacific_Islander / Total,
    'Other' = Other / Total,
    'Multi-Racial' = `Multi-Racial` / Total,
    'name' = name
  ) %>% filter(!is.na(.))
# this gives us census tracts with race breakdowns by percentage
```











new (post v2):
```{r SELECTED HOUSING CHARACTERISTICS DATA}
sel_house_chars_2013 <- read_csv(
  "data/selected_housing_characteristics/ACSDP5Y2013.DP04_data_with_overlays_2021-12-12T194948.csv")

    # 61, 65, 69, 73, 77, 81, 85, 89, 93, 97,   # year structure built
    # 21, 25, 29,   # units in structure
    # 173, 177, 181, 185, 189,   # housing tenure
    # 193, 197, 201, 205, 209, 213, 217,   # year householder moved in
    # 313, 317, 321, 325, 329, 333, 337, 341, 345, 349,   # Value
    # 565   # GEO_ID

units_in_structure_2013 <- sel_house_chars_2013 %>% 
  select(21, 25, 29, 565) %>% 
  row_to_names(row_number = 1) %>% 
  transmute(
    "Total housing units" = as.numeric(`Estimate!!UNITS IN STRUCTURE!!Total housing units`),
    "1-unit, detached" = as.numeric(`Estimate!!UNITS IN STRUCTURE!!Total housing units!!1-unit, detached`),
    "1-unit, attached" = as.numeric(`Estimate!!UNITS IN STRUCTURE!!Total housing units!!1-unit, attached`),
    name = substr(id, 10, 20)               
  ) %>% 
  transmute(
    "total_1_units" = (`1-unit, detached` + `1-unit, attached`),
    "perc_1_unit" = (`1-unit, detached` + `1-unit, attached`) / `Total housing units`,
    name = name
  )
  
units_in_structure_2014 <- sel_house_chars_2014 %>% 
  select(21, 25, 29, 565) %>% 
  row_to_names(row_number = 1) %>% 
  transmute(
    "Total housing units" = as.numeric(`Estimate!!UNITS IN STRUCTURE!!Total housing units`),
    "1-unit, detached" = as.numeric(`Estimate!!UNITS IN STRUCTURE!!Total housing units!!1-unit, detached`),
    "1-unit, attached" = as.numeric(`Estimate!!UNITS IN STRUCTURE!!Total housing units!!1-unit, attached`),
    name = substr(id, 10, 20)               
  ) %>% 
  transmute(
    "single_family_perc" = (`1-unit, detached` + `1-unit, attached`) / `Total housing units`,
    name = name
  )

##

housing_tenure_2013 <- sel_house_chars_2013 %>% 
  select(173, 177, 181, 185, 189) %>% 
  row_to_names(row_number = 1) %>% 
  transmute(
    "Occupied Housing Units" = `Estimate!!HOUSING TENURE!!Occupied housing units`,
    "Owner-occupied" = `Estimate!!HOUSING TENURE!!Occupied housing units!!Owner-occupied`,
    "Renter-Occupied" = `Estimate!!HOUSING TENURE!!Occupied housing units!!Renter-occupied`,
    "Average household size of owner-occupied unit" = `Estimate!!HOUSING TENURE!!Average household size of owner-occupied unit`,
    "Average household size of renter-occupied unit" = `Estimate!!HOUSING TENURE!!Average household size of renter-occupied unit`
  )


  

sel_house_chars_2014 <- read_csv(
  "data/selected_housing_characteristics/ACSDP5Y2014.DP04_data_with_overlays_2021-12-12T194948.csv")

sel_house_chars_2015 <- read_csv(
  "data/selected_housing_characteristics/ACSDP5Y2015.DP04_data_with_overlays_2021-12-12T194948.csv")

sel_house_chars_2016 <- read_csv(
  "data/selected_housing_characteristics/ACSDP5Y2016.DP04_data_with_overlays_2021-12-12T194948.csv")

sel_house_chars_2017 <- read_csv(
  "data/selected_housing_characteristics/ACSDP5Y2017.DP04_data_with_overlays_2021-12-12T194948.csv")

bay_tracts_need <- readRDS("")

# build and filter 2013, then make loop to put all years into one
# can then filter more efficiently (fewer lines of code)

years <- 2013:2017

sel_house_chars <- NULL

for(year in years) {
  
      filename <-
        paste0(
          "data/selected_housing_characteristics/ACSDP5Y",
          year,
          ".DP04_data_with_overlays_2021-12-12T194948.csv"
        )

      print(filename)
      
      temp <- read_csv(filename)

      sel_house_chars <-
        rbind(sel_house_chars, temp)
}
```




































