---
title: "Energy Efficient Homes in the Bay Area: A Data-Driven Look into Geography and Sociodemographic Trends"
author: "Swen Hendrickson"
date: "12/14/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r 1. setup, include = F}
knitr::opts_chunk$set(echo = F, warning = F, message = F)

# Clear all variables, close all plots, and clear the console, load libraries
# rm(list=ls())
# graphics.off()
# cat("\014")

setwd("~/GitHub/swehen17.github.io/218X/Project")

library(censusapi)
library(tidyverse)
library(tigris)
library(sf)
library(leaflet)
library(mapview)
library(plotly)
library(data.table)
library(viridis)
library(hrbrthemes)
library(knitr)
library(flexdashboard)

Sys.setenv(CENSUS_KEY="c8aa67e4086b4b5ce3a8717f59faa9a28f611dab")

acs_vars_2019_5yr <- readRDS("acs_vars_2019_5yr.rds")
```

# Introduction

## Abstract
The driving question behind this project is "WHERE are energy efficient homes in the Bay Area and WHO benefits from occupying them?" Behind this question is a foundational understanding that occupying an energy efficient home has financial benefits. The most obvious advantage is that utility bills are lower than they otherwise would be, resulting in cost savings that accumulate over time. It has also been documented that energy efficient homes earn a greater resale value (LINK?). Achieving energy efficiency, however, costs money, either through retrofits of an old home or brand new construction (which is increasingly rare in the highly developed Bay Area). Due to this barrier to entry, it would logically follow that residential energy efficiency is one of the countless mechanisms in our society that perpetuates economic disparity. The following analysis seeks to view this issue from a number of angles, picking it apart with statistical analyses and placing on the map what the energy consumption landscape looks like in the Bay Area. 

## Data
Energy consumption data comes from aggregated PG&E data, via another data project called "The Bay Area Energy Atlas" from the organization BayREN (LINK?). The data file I used included a combined annual electricity and natural gas usage figure for each year from 2013 to 2017, aggregated to the Census Tract level but split into categories of building type (residential, industrial, etc.). For my analysis, I primarily used the residential sub-category: single-family homes. In addition to the combined usage figure, numbers on population and 'built square feet' for each building type allowed for calculation of varying consumption metrics (detailed below). 

Socio-demographic and housing characteristics data for cross-analysis was mostly brought in from ACS 5-year estimate surveys, which allowed for selection of individual years in order to match the time frame of the PG&E data. In the case of race (used for equity analysis) data came from the 2010 census. 

(Note: earlier on in the project, an attempt was made to incorporate all five years of data and build dashboards with year-selection menus, so that maps and plots could be created for each year. In the end, however, due to time constraints, analysis was only performed for the year 2013).

## Defining Energy Consumption: 3 Metrics
### EISF - Energy Intensity per Square Foot
Energy Intensity is a standard measure of building energy efficiency which is typically expressed in Btu/SF or kBTu/SF. Because it is normalized by space size, it allows for comparison between buildings of varying sizes. Below is a map showing averaged EISF values for Bay Area tracts. A darker orange means higher energy intensity and on average less efficient buildings, and vice versa.

```{r MAP: EISF}
bay_eng_int_map = readRDS("bay_eng_int_map.rds")
beim_2013 = read_rds("beim_2013.rds")

# actually mapping, finally
# 2013
beim_pal <- colorNumeric(
  palette = "Oranges",
  domain = bay_eng_int_map$energy_intensity
)

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    data = beim_2013,
    fillColor = ~beim_pal(energy_intensity),
    color = "white",
    weight = 0.5,
    fillOpacity = 0.5,
    label = ~energy_intensity
  )
```

### ECPC - Energy Consumption Per Capita
Instead of measuring energy consumption at the building level (for which Energy Intensity is the default measure), considering per person usage could offer unique insights. The PG&E data gathered through BayREN contained a 'Usage Per Capita' column for all residential buildings, allowing for the derivation of the ECPC measure. (Note: ECPC was derived using the proportionality of single-family homes to total residential buildings by built square feet. ACS and Census data was searched at length for a population figure of people living in 'single-family homes' to no avail). The map below shows varying levels of ECPC for Census tracts across the Bay Area.

```{r MAP: ECPC}
ECPC_2013 <- readRDS("ECPC_2013.rds")

ECPC_pal <- colorNumeric(
  palette = "Oranges",
  domain = ECPC_2013$ecpc
)

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    data = ECPC_2013,
    fillColor = ~ECPC_pal(ecpc),
    color = "white",
    weight = 0.5,
    fillOpacity = 0.5,
    label = ~ecpc
  )
```

### ECPH - Energy Consumption Per Household
For this metric, ACS data was collected (from table DP04: "Selected Housing Characteristics") that provided the number of "1-unit" households in each tract. For purposes of this analysis, "1-unit" households are equated to "single-family" homes. Thus, the single-family energy consumption number from the PG&E data was divided over the ACS data on number of 1-unit households per tract to arrive at this metric for household consumption, averaged across each tract. ECPH is similarly mapped below for 2013.

```{r MAP: ECPH}
ECPH_2013 <- readRDS("ECPH_2013.rds")

ECPH_pal <- colorNumeric(
  palette = "Oranges",
  domain = ECPH_2013$ecph
)

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    data = ECPH_2013,
    fillColor = ~ECPH_pal(ecph),
    color = "white",
    weight = 0.5,
    fillOpacity = 0.5,
    label = ~ecph
  )
```

# Analysis
Having established the three metrics for energy consumption and efficiency above (EISF, ECPC, and ECPH), this data was cross-examined with selected socio-demographic and housing characteristics which might have links with energy use. Simple linear regression was used throughout as a useful method for assessing the strength of these possible relationships. On the social side, the two variables examined below are income and race. Takeaways from these analyses are relevant to discussions of equity, with the understanding described above that energy efficiency perpetuates financial disparities between groups. On the housing side, three variables were assessed: tenure statues (whether an occupant owns or rents), the number of years spent living in a home, and the value of a home. These variables (five in total), came from ACS and Census data, and with the exception of race, match the year of the energy data used (2013). As a reminder, data was necessarily collected at the tract level to match the smallest possible geographic granularity offered for energy consumption.

## Income
A weighted average income figure was calculated for each tract by multiplying the number of people in each income bracket by the midpoint income of that bracket.

### EISF vs. Income
```{r ANALYSIS: EISF vs. INCOME}
bay_tracts_use_geo = readRDS("bay_tracts_use_geo.rds")

avg_house_size_2013 = readRDS("avg_house_size_2013.rds")

avg_house_size_2013_map <- avg_house_size_2013 %>% 
  left_join(bay_tracts_use_geo, by = "name") %>% 
  st_as_sf()

# where are the tracts with very large houses?
# top_10_house_size_tracts <- avg_house_size_2013_map %>% 
#   filter(avg_house_size > 5755)
# 
# mapview(top_10_house_size_tracts)

# avg_house_size_2013_map_PAL <- colorNumeric(
#   palette = "Oranges",
#   domain = avg_house_size_2013_map$avg_house_size
# )

# leaflet() %>%
#   addProviderTiles(providers$CartoDB.Positron) %>%
#   addPolygons(
#     data = avg_house_size_2013_map,
#     fillColor = ~avg_house_size_2013_map_PAL(avg_house_size),
#     color = "white",
#     weight = 0.5,
#     fillOpacity = 0.5,
#     label = ~avg_house_size
#   )

# plot energy intensity by income
# 2013
income_2013 = readRDS("income_2013.rds")

income_engint_2013 <- beim_2013 %>%
  left_join(income_2013, by = "name") %>%
  select(weighted_avg_income, energy_intensity, name) %>%
  st_drop_geometry() %>% 
  left_join(avg_house_size_2013, by = "name")

ggplot(
    data = income_engint_2013,
    aes(
      x = weighted_avg_income,
      y = energy_intensity
    )
  ) +
  geom_point(
    aes(
      color = avg_house_size
    )
  ) +
  ggtitle("Energy Intensity vs. Income") +
  xlab("Weighted Average Household Income (USD)") +
  ylab("Energy Intensity (Btu/SF)") +
  geom_smooth(method = "lm")
```

We see in the plot above that, as suspected, income has a negative association with energy intensity. In other words, as income goes up, people's homes tend to get more energy efficient. House size is also incorporated here and displayed by color, with tracts having on average larger homes showing up in lighter shades of blue. Also as expected, house size increases with increased income.

There seems to be a low cluster of more efficient tracts separated from the pack..

```{r ANALYSIS: EISF vs. INCOME 2}
# color code low EISF cluster
cluster <- income_engint_2013 %>%
  filter(energy_intensity < 25000)

ggplot(
    data = income_engint_2013,
    aes(
      x = weighted_avg_income,
      y = energy_intensity
    )
  ) +
  geom_point() +
  ggtitle("Energy Intensity vs. Income") +
  xlab("Weighted Average Household Income (USD)") +
  ylab("Energy Intensity (Btu/SF)") +
  # geom_smooth(method = "lm") +
  geom_point(data = cluster, color = "purple")

# TO UPDATE TITLE FORMATTING:
# theme_update(plot.title = element_text(size = 22))
# theme(plot.title = element_text(hjust = 0.5)) +
#

# as income goes up, energy intensity per SF goes down
# cluster of low intensity bldgs
# TODO: look into where these buildings are

model <- lm(energy_intensity ~ weighted_avg_income, data = income_engint_2013)

# summary(model)
```

Where are those more efficient tracts?

```{r ANALYSIS: EISF vs. INCOME 3}
# where are the super efficient tracts?
efficient_tracts_2013 <- beim_2013 %>%
  filter(energy_intensity < 25000)

# map it
beim_pal <- colorNumeric(
  palette = "Green",
  domain = efficient_tracts_2013$energy_intensity
)

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    data = efficient_tracts_2013,
    fillColor = ~beim_pal(energy_intensity),
    color = "white",
    weight = 0.5,
    fillOpacity = 0.5,
    label = ~energy_intensity
  )
# noticeably, big urban centers are missing (SF, Oak, San Jose)
# mostly less populous tracts, but peninsula oddly well-represented
# (Daly City, South San Francisco, San Mateo, San Mateo County)
# theories: 
# for less populated tracts: weather patterns, weekend homes, vacation homes (i.e. vacant more often)
# for the peninsula: abundance of Silicon Valley workers who moved in relatively recently, have income, are educated on energy efficiency, and have built or upgraded homes recently

# explore deeper
# of the 150 low EISF tracts above, break into 3 groups, by EISF
# efficient_tracts_2013_tier1 <- efficient_tracts_2013 %>% 
#   filter(energy_intensity < 19351.275)
# mapview(efficient_tracts_2013_tier1)
# 
# efficient_tracts_2013_tier2 <- efficient_tracts_2013 %>% 
#   filter(energy_intensity > 19351.275 & energy_intensity < 21552.34)
# mapview(efficient_tracts_2013_tier2)
# 
# efficient_tracts_2013_tier3 <- efficient_tracts_2013 %>% 
#   filter(energy_intensity > 21552.34)
# mapview(efficient_tracts_2013_tier3)
```

Noticeably, big urban centers are missing from this map (San Francisco, Oakland, San Jose). The map seems to be made up of less populated tracts with more open space, and towns on the Peninsula (San Mateo, South San Francisco, etc.). The lower EISF numbers in the more "countryside" tracts could be a result of homes that are not regularly used and thus sit vacant much of the time. Also note, for sparsely populated tracts where the base of data observations is much smaller, efficiency averages can be more easily influenced by a smaller number of buildings. The low EISF tracts on the Peninsula, in the corridor leading to Silicon Valley could be a product of tech-oriented residents who work in the technology industry and likely have the means and education to invest in home efficiency measures. These are just theories for many possible reasons leading to this cluster.

### ECPC vs. Income

```{r ANALYSIS: ECPC vs. INCOME}
# plot ECPC by income
ECPC = readRDS("ECPC.rds")

# 2013
income_ECPC_2013 <- ECPC %>%
  filter(year == 2013, !is.na(usage_percap)) %>%
  left_join(income_2013, by = "name") %>%
  select(weighted_avg_income, usage_percap)

ggplot(
    data = income_ECPC_2013,
    aes(
      x = weighted_avg_income,
      y = usage_percap
    )
  ) +
  geom_point() +
  ggtitle("Energy Consumption Per Capita vs. Income") +
  xlab("Weighted Average Household Income (USD)") +
  ylab("Energy Consumption Per Capita (Btu/Person)") +
  geom_smooth(method = "lm")

# as income goes up, consumption per capita increases steadily!

model <- lm(usage_percap ~ weighted_avg_income, data = income_ECPC_2013)

# summary(model)
```

When we change our consumption metric from Energy Intensity to Per Capita consumption (ECPC), we see that on an individual level, people tend to consume more as their income goes up. There is an interesting relationship here, because we saw above that as income goes up, HOUSES tend to be more efficient (per SF). The confounding variable leading to the opposite slopes of these two graphs is almost certainly house size, in that EISF numbers will shrink as income rises and houses become more spacious (as the denomenator of square feet increases). This, in combination with the fact that efficient homes (which cost more money) will be occupied by higher-income inhabitants, allows ECPC to rise as income rises while EISF falls.

## Equity Analysis
   Equity analysis in this case is tricky, since consumption data from the utility (PG&E) was, for privacy reasons, aggregated to the census tract level as the smallest geographical granularity. This means that we can compare a few census tracts by their overall efficiency scores (EISF for single family homes) and their overall racial makeup, but we cannot say for certain which specific houses operate more or less efficiently and who exactly lives in those houses.
  So, the tracts with both the highest and the lowest EISF scores (shown in the EISF maps) are inspected for their racial make up to see if any trends are apparent.
   
```{r ANALYSIS: EQUITY ANALYSIS}
# find best and worst tracts by EISF (using 2013-2017 avg)
 # data: beim_2013to2013_avgs
beim_2013to2017_avgs = readRDS("beim_2013to2017_avgs.rds")
# then join with race_2010 by name to produce pie charts
race_2010 = readRDS("race_2010.rds")

# MOST energy efficient tract:
lowest_EISF_tract_2013to2017 <- beim_2013to2017_avgs %>% 
  filter(name == "06081613501") %>% 
  left_join(race_2010, by = "name")



# join w/ race data
data1 <- lowest_EISF_tract_2013to2017[ , 4:10] %>% st_drop_geometry()
# plot
data <- data.frame(
  name = c(colnames(data1)) ,  
  value = c(as.numeric(data1[1, ]))
  )
# ..
ggplot(data, aes(x = name, y = value, fill = name)) + 
  geom_bar(stat = "identity") +
  ggtitle("Tract with LOWEST EISF") +
  xlab("Race Category") +
  ylab("Percentage of Population")

# map
mapview(lowest_EISF_tract_2013to2017)

# LEST efficient tract:
highest_EISF_tract_2013to2017 <- beim_2013to2017_avgs %>% 
  filter(name == "06013314104") %>% 
  left_join(race_2010, by = "name")

# join w/ race data
data1 <- highest_EISF_tract_2013to2017[ , 4:10] %>% st_drop_geometry()
# plot
data <- data.frame(
  name = c(colnames(data1)) ,  
  value = c(as.numeric(data1[1, ]))
  )
# ..
ggplot(data, aes(x = name, y = value, fill = name)) + 
  geom_bar(stat = "identity") +
  ggtitle("Tract with HIGHEST EISF") +
  xlab("Race Category") +
  ylab("Percentage of Population")

# map
mapview(highest_EISF_tract_2013to2017)

# TODO: could also find Q1 and Q3 (circle back later)
```

As we can see above, the tract with the most efficient homes (located in a hilly and coastal region of San Mateo county) is much less diverse than the tract with the highest EISF (located in Pittsburg). The sample size here is small, but the implications for energy efficiency and equity speak for themselves. 


```{r ANALYSIS: ECPH & ECPC vs. TENURE}
housing_tenure_2013 <- readRDS("housing_tenure_2013.rds")

# ECPH vs. TENURE
ECPH_vs_TENURE_2013 <- ECPH_2013 %>% 
  left_join(housing_tenure_2013, by = "name") %>% 
  st_drop_geometry() %>% 
  select(c(ecph, owned_units_perc))

ggplot(
    data = ECPH_vs_TENURE_2013,
    aes(
      x = owned_units_perc,
      y = ecph
    )
  ) +
  geom_point() +
  ggtitle("Energy Consumption Per Houseold vs. Percent of Units OWNED") +
  xlab("Percentage of Units in Tract that are Owned") +
  ylab("Average Energy Consumption Per Household (Btu/Unit)") +
  geom_smooth(method = "lm")

# seem to have a positive association between ownership and consumption

model <- lm(ecph ~ owned_units_perc, data = ECPH_vs_TENURE_2013)

summary(model)

##

ECPC_2013 <- readRDS("ECPC_2013.rds")

# ECPC vs. TENURE
ECPC_vs_TENURE_2013 <- ECPC_2013 %>% 
  left_join(housing_tenure_2013, by = "name") %>% 
  select(c(ecpc, owned_occ_perc))

ggplot(
    data = ECPC_vs_TENURE_2013,
    aes(
      x = owned_occ_perc,
      y = ecpc
    )
  ) +
  geom_point() +
  ggtitle("Energy Consumption Per Capita vs. Percent of Occupants in OWNED units") +
  xlab("Percentage of Occupants in Tract in Owned Units") +
  ylab("Average Energy Consumption Per Capita (Btu/Person)") +
  geom_smooth(method = "lm")

# again, ownership seems to lead to more consumption

model <- lm(ecpc ~ owned_occ_perc, data = ECPC_vs_TENURE_2013)

summary(model)
```
The above (2) plots show that increased ownership is associated with increased consumption, both when measured per household and per occupant

```{r ANALYSIS: EISF vs. TENURE}
EISF_vs_TENURE_2013 <- beim_2013 %>% 
  st_drop_geometry() %>% 
  left_join(housing_tenure_2013, by = "name") %>% 
  select(energy_intensity, owned_units_perc)

ggplot(
    data = EISF_vs_TENURE_2013,
    aes(
      x = owned_units_perc,
      y = energy_intensity
    )
  ) +
  geom_point() +
  ggtitle("Energy Intensity vs. Ownership") +
  xlab("Percentage of Units in Tract that are Owned") +
  ylab("Energy Intensity for Homes in Tract (Btu/SF)") +
  geom_smooth(method = "lm")

# unclear or flat relationship between ownership and EISF

model <- lm(energy_intensity ~ owned_units_perc, data = EISF_vs_TENURE_2013)

summary(model)

# slope almost flat, R^2 low (0.07)
```

```{r ANLYSIS: ECPH vs. YEARS IN HOME}
years_in_home_2013 <- readRDS("years_in_home_2013.rds")

ECPH_vs_YEARS_in_home_2013 <- ECPH_2013 %>% 
  st_drop_geometry() %>% 
  left_join(years_in_home_2013, by = "name") %>% 
  select(ecph, weighted_avg_years_in_home)

ggplot(
    data = ECPH_vs_YEARS_in_home_2013,
    aes(
      x = weighted_avg_years_in_home,
      y = ecph
    )
  ) +
  geom_point() +
  ggtitle("Energy Consumption (per Household) vs. YEARS Spent In Home") +
  xlab("Weighted Average Number of Years Spent in Home") +
  ylab("Average Energy Consumption (per Household) for Tract (Btu/unit)") +
  geom_smooth(method = "lm")

# implies that consumption is higher for residents who have lived in their homes for longer
# likely influence (theory): that more efficient homes are occupied by people who have more recently moved into the Bay

model <- lm(ecph ~ weighted_avg_years_in_home, data = ECPH_vs_YEARS_in_home_2013)

summary(model)

# R^2 (0.27)
```

```{r ANLYSIS: ECPH vs. VALUE}
VALUE_2013 <- readRDS("VALUE_2013.rds")

# consumption per household vs. home values (of owner-occupied units)
ECPH_vs_VALUE_2013 <- ECPH_2013 %>% 
  st_drop_geometry() %>% 
  left_join(VALUE_2013, by = "name")

ggplot(
    data = ECPH_vs_VALUE_2013,
    aes(
      x = weighted_avg_value,
      y = ecph
    )
  ) +
  geom_point() +
  ggtitle("Energy Consumption (per Household) vs. HOME VALUE") +
  xlab("Weighted Average Home Value for Owner-Occupied Units in Tract") +
  ylab("Average Energy Consumption (per Household) for Tract (Btu/unit)") +
  geom_smooth(method = "lm")

# seems to be positive but mild association bw home value and consumption

model <- lm(ecph ~ weighted_avg_value, data = ECPH_vs_VALUE_2013)

summary(model)

# slope almost low, R^2 low (0.22)
```

# Conclusion
In conclusion, the main driving hypothesis of this analysis seems to be confirmed: that energy efficient homes tend to be owned and occupied by more members of higher income brackets. An interesting related trend was also revealed: that higher earners also consume more on a per capita basis. So by one view (using EISF), we might applaud the efforts of wealthy homeowners who make their homes more energy efficient. At the same time, however, wealthier indviduals appear to consume more per capita (measured by ECPC). This consumption appears to be masked by the efficiency of the homes in question, as well as the size of the homes in question (as a larger but equally consuming building will have a lower EISF). Our equity analysis also shed some light on the racial makeup of the most extreme tracts in the analysis (minimum and maximum EISF); the tract with the highest efficiency was predominantly White with some Asian population, while the tract with the lowest EISF scores was much more racially diverse. Both conclusions point to some of the disparities surrounding residential energy efficiency. And while there should certainly be a goal to 'democratize' building energy efficiency for environmental reasons, we should also examine our apparent tendency to consume more energy as we earn more and can afford more efficient homes. 























