---
title: "HendricksonSwen_Project"
author: "Swen Hendrickson"
date: "12/1/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r 1. setup, include = F}
knitr::opts_chunk$set(echo = F, warning = F, message = F)

# Clear all variables, close all plots, and clear the console, load libraries
# rm(list=ls())
# graphics.off()
# cat("\014")

setwd("~/GitHub/swehen17.github.io/218X/Project")

library(censusapi)
library(tidyverse)
library(tigris)
library(sf)
library(leaflet)
library(mapview)
library(plotly)
library(data.table)
library(viridis)
library(hrbrthemes)
library(knitr)
library(flexdashboard)

Sys.setenv(CENSUS_KEY="c8aa67e4086b4b5ce3a8717f59faa9a28f611dab")

acs_vars_2019_5yr <- readRDS("acs_vars_2019_5yr.rds")
```

```{r MAP: EISF}
include_url("https://swehen17.shinyapps.io/HendricksonSwen_Project_DASHBOARD/?_ga=2.211101141.1564337814.1639522713-1483363889.1637271700")

include_app("https://stanfordfuturebay.shinyapps.io/dashboard_demo2/", height = "800px")

include_url("https://stanfordfuturebay.shinyapps.io/dashboard_demo2/", height = "1200px")

#

bay_eng_int_map = readRDS("bay_eng_int_map.rds")
beim_2013 = read_rds("beim_2013.rds")

# actually mapping, finally
# 2013
beim_pal <- colorNumeric(
  palette = "Oranges",
  domain = bay_eng_int_map$energy_intensity
)

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    data = beim_2013,
    fillColor = ~beim_pal(energy_intensity),
    color = "white",
    weight = 0.5,
    fillOpacity = 0.5,
    label = ~energy_intensity
  )
```

```{r MAP: ECPC}
ECPC_2013 <- readRDS("ECPC_2013.rds")

ECPC_pal <- colorNumeric(
  palette = "Oranges",
  domain = ECPC_2013$ecpc
)

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    data = ECPC_2013,
    fillColor = ~ECPC_pal(ecpc),
    color = "white",
    weight = 0.5,
    fillOpacity = 0.5,
    label = ~ecpc
  )
```
Methodology of above:
- only had population for total residential (includes mutli family, etc)
- but had SF data for single family homes. got percentage that single family
homes make up of total residential, to multiply by the usage per capita for
total residential to get usage per capita for single family homes
- helps to control for neighborhoods that might have different proportions 
of homes vs. multifamily complexes, mobile homes, etc MAP: ECPH.

```{r MAP: ECPH}
ECPH_2013 <- readRDS("ECPH_2013.rds")

ECPH_pal <- colorNumeric(
  palette = "Oranges",
  domain = ECPH_2013$ecph
)

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    data = ECPH_2013,
    fillColor = ~ECPH_pal(ecph),
    color = "white",
    weight = 0.5,
    fillOpacity = 0.5,
    label = ~ecph
  )
```



```{r ANALYSIS: EISF vs. INCOME}
bay_tracts_use_geo = readRDS("bay_tracts_use_geo.rds")

avg_house_size_2013 = readRDS("avg_house_size_2013.rds")

avg_house_size_2013_map <- avg_house_size_2013 %>% 
  left_join(bay_tracts_use_geo, by = "name") %>% 
  st_as_sf()

# where are the tracts with very large houses?
# top_10_house_size_tracts <- avg_house_size_2013_map %>% 
#   filter(avg_house_size > 5755)
# 
# mapview(top_10_house_size_tracts)

avg_house_size_2013_map_PAL <- colorNumeric(
  palette = "Oranges",
  domain = avg_house_size_2013_map$avg_house_size
)

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    data = avg_house_size_2013_map,
    fillColor = ~avg_house_size_2013_map_PAL(avg_house_size),
    color = "white",
    weight = 0.5,
    fillOpacity = 0.5,
    label = ~avg_house_size
  )

# plot energy intensity by income
# 2013
income_2013 = readRDS("income_2013.rds")

income_engint_2013 <- beim_2013 %>%
  left_join(income_2013, by = "name") %>%
  select(weighted_avg_income, energy_intensity, name) %>%
  st_drop_geometry() %>% 
  left_join(avg_house_size_2013, by = "name")

ggplot(
    data = income_engint_2013,
    aes(
      x = weighted_avg_income,
      y = energy_intensity
    )
  ) +
  geom_point(
    aes(
      color = avg_house_size
    )
  ) +
  ggtitle("Energy Intensity vs. Income") +
  xlab("Weighted Average Household Income (USD)") +
  ylab("Energy Intensity (Btu/SF)") +
  geom_smooth(method = "lm")

# color code low EISF cluster
cluster <- income_engint_2013 %>%
  filter(energy_intensity < 25000)

ggplot(
    data = income_engint_2013,
    aes(
      x = weighted_avg_income,
      y = energy_intensity
    )
  ) +
  geom_point() +
  ggtitle("Energy Intensity vs. Income") +
  xlab("Weighted Average Household Income (USD)") +
  ylab("Energy Intensity (Btu/SF)") +
  # geom_smooth(method = "lm") +
  geom_point(data = cluster, color = "purple")

# TO UPDATE TITLE FORMATTING:
# theme_update(plot.title = element_text(size = 22))
# theme(plot.title = element_text(hjust = 0.5)) +
#

# as income goes up, energy intensity per SF goes down
# cluster of low intensity bldgs
# TODO: look into where these buildings are

model <- lm(energy_intensity ~ weighted_avg_income, data = income_engint_2013)

summary(model)

# where are the super efficient tracts?
efficient_tracts_2013 <- beim_2013 %>%
  filter(energy_intensity < 25000)

# map it
beim_pal <- colorNumeric(
  palette = "Green",
  domain = efficient_tracts_2013$energy_intensity
)

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    data = efficient_tracts_2013,
    fillColor = ~beim_pal(energy_intensity),
    color = "white",
    weight = 0.5,
    fillOpacity = 0.5,
    label = ~energy_intensity
  )
# noticeably, big urban centers are missing (SF, Oak, San Jose)
# mostly less populous tracts, but peninsula oddly well-represented
# (Daly City, South San Francisco, San Mateo, San Mateo County)
# theories: 
# for less populated tracts: weather patterns, weekend homes, vacation homes (i.e. vacant more often)
# for the peninsula: abundance of Silicon Valley workers who moved in relatively recently, have income, are educated on energy efficiency, and have built or upgraded homes recently

# explore deeper
# of the 150 low EISF tracts above, break into 3 groups, by EISF
efficient_tracts_2013_tier1 <- efficient_tracts_2013 %>% 
  filter(energy_intensity < 19351.275)
mapview(efficient_tracts_2013_tier1)

efficient_tracts_2013_tier2 <- efficient_tracts_2013 %>% 
  filter(energy_intensity > 19351.275 & energy_intensity < 21552.34)
mapview(efficient_tracts_2013_tier2)

efficient_tracts_2013_tier3 <- efficient_tracts_2013 %>% 
  filter(energy_intensity > 21552.34)
mapview(efficient_tracts_2013_tier3)
```

```{r ANALYSIS: ECPC vs. INCOME}
# plot ECPC by income
ECPC = readRDS("ECPC.rds")

# 2013
income_ECPC_2013 <- ECPC %>%
  filter(year == 2013, !is.na(usage_percap)) %>%
  left_join(income_2013, by = "name") %>%
  select(weighted_avg_income, usage_percap)

ggplot(
    data = income_ECPC_2013,
    aes(
      x = weighted_avg_income,
      y = usage_percap
    )
  ) +
  geom_point() +
  ggtitle("Energy Consumption Per Capita vs. Income") +
  xlab("Weighted Average Household Income (USD)") +
  ylab("Energy Consumption Per Capita (Btu/Person)") +
  geom_smooth(method = "lm")

# as income goes up, consumption per capita increases steadily!

model <- lm(usage_percap ~ weighted_avg_income, data = income_ECPC_2013)

summary(model)
```

```{r ANALYSIS: ECPH & ECPC vs. TENURE}
housing_tenure_2013 <- readRDS("housing_tenure_2013.rds")

# ECPH vs. TENURE
ECPH_vs_TENURE_2013 <- ECPH_2013 %>% 
  left_join(housing_tenure_2013, by = "name") %>% 
  st_drop_geometry() %>% 
  select(c(ecph, owned_units_perc))

ggplot(
    data = ECPH_vs_TENURE_2013,
    aes(
      x = owned_units_perc,
      y = ecph
    )
  ) +
  geom_point() +
  ggtitle("Energy Consumption Per Houseold vs. Percent of Units OWNED") +
  xlab("Percentage of Units in Tract that are Owned") +
  ylab("Average Energy Consumption Per Household (Btu/Unit)") +
  geom_smooth(method = "lm")

# seem to have a positive association between ownership and consumption

model <- lm(ecph ~ owned_units_perc, data = ECPH_vs_TENURE_2013)

summary(model)

##

ECPC_2013 <- readRDS("ECPC_2013.rds")

# ECPC vs. TENURE
ECPC_vs_TENURE_2013 <- ECPC_2013 %>% 
  left_join(housing_tenure_2013, by = "name") %>% 
  select(c(ecpc, owned_occ_perc))

ggplot(
    data = ECPC_vs_TENURE_2013,
    aes(
      x = owned_occ_perc,
      y = ecpc
    )
  ) +
  geom_point() +
  ggtitle("Energy Consumption Per Capita vs. Percent of Occupants in OWNED units") +
  xlab("Percentage of Occupants in Tract in Owned Units") +
  ylab("Average Energy Consumption Per Capita (Btu/Person)") +
  geom_smooth(method = "lm")

# again, ownership seems to lead to more consumption

model <- lm(ecpc ~ owned_occ_perc, data = ECPC_vs_TENURE_2013)

summary(model)
```
The above (2) plots show that increased ownership is associated with increased consumption, both when measured per household and per occupant

```{r ANALYSIS: EISF vs. TENURE}
EISF_vs_TENURE_2013 <- beim_2013 %>% 
  st_drop_geometry() %>% 
  left_join(housing_tenure_2013, by = "name") %>% 
  select(energy_intensity, owned_units_perc)

ggplot(
    data = EISF_vs_TENURE_2013,
    aes(
      x = owned_units_perc,
      y = energy_intensity
    )
  ) +
  geom_point() +
  ggtitle("Energy Intensity vs. Ownership") +
  xlab("Percentage of Units in Tract that are Owned") +
  ylab("Energy Intensity for Homes in Tract (Btu/SF)") +
  geom_smooth(method = "lm")

# unclear or flat relationship between ownership and EISF

model <- lm(energy_intensity ~ owned_units_perc, data = EISF_vs_TENURE_2013)

summary(model)

# slope almost flat, R^2 low (0.07)
```

```{r ANLYSIS: ECPH vs. VALUE}
VALUE_2013 <- readRDS("VALUE_2013.rds")

# consumption per household vs. home values (of owner-occupied units)
ECPH_vs_VALUE_2013 <- ECPH_2013 %>% 
  st_drop_geometry() %>% 
  left_join(VALUE_2013, by = "name")

ggplot(
    data = ECPH_vs_VALUE_2013,
    aes(
      x = weighted_avg_value,
      y = ecph
    )
  ) +
  geom_point() +
  ggtitle("Energy Consumption (per Household) vs. HOME VALUE") +
  xlab("Weighted Average Home Value for Owner-Occupied Units in Tract") +
  ylab("Average Energy Consumption (per Household) for Tract (Btu/unit)") +
  geom_smooth(method = "lm")

# seems to be positive but mild association bw home value and consumption

model <- lm(ecph ~ weighted_avg_value, data = ECPH_vs_VALUE_2013)

summary(model)

# slope almost low, R^2 low (0.22)
```

```{r ANLYSIS: ECPH vs. YEARS IN HOME}
years_in_home_2013 <- readRDS("years_in_home_2013.rds")

ECPH_vs_YEARS_in_home_2013 <- ECPH_2013 %>% 
  st_drop_geometry() %>% 
  left_join(years_in_home_2013, by = "name") %>% 
  select(ecph, weighted_avg_years_in_home)

ggplot(
    data = ECPH_vs_YEARS_in_home_2013,
    aes(
      x = weighted_avg_years_in_home,
      y = ecph
    )
  ) +
  geom_point() +
  ggtitle("Energy Consumption (per Household) vs. YEARS Spent In Home") +
  xlab("Weighted Average Number of Years Spent in Home") +
  ylab("Average Energy Consumption (per Household) for Tract (Btu/unit)") +
  geom_smooth(method = "lm")

# implies that consumption is higher for residents who have lived in their homes for longer
# likely influence (theory): that more efficient homes are occupied by people who have more recently moved into the Bay

model <- lm(ecph ~ weighted_avg_years_in_home, data = ECPH_vs_YEARS_in_home_2013)

summary(model)

# R^2 (0.27)
```




# Equity Analysis
   Equity analysis in this case is tricky, since consumption data from the utility (PG&E) was, for privacy reasons, aggregated to the census tract level as the smallest geographical granularity. This means that we can compare a few census tracts by their overall efficiency scores (EISF for single family homes) and their overall racial makeup, but we cannot say for certain which specific houses operate more or less efficiently and who exactly lives in those houses.
  So, the tracts with both the highest and the lowest EISF scores (shown in the EISF maps) will be inspected for their racial make up to see if any trends are apparent.
   
```{r ANALYSIS: EQUITY ANALYSIS}
# find best and worst tracts by EISF (using 2013-2017 avg)
 # data: beim_2013to2013_avgs
beim_2013to2017_avgs = readRDS("beim_2013to2017_avgs.rds")
# then join with race_2010 by name to produce pie charts
race_2010 = readRDS("race_2010.rds")

# MOST energy efficient tract:
lowest_EISF_tract_2013to2017 <- beim_2013to2017_avgs %>% 
  filter(name == "06081613501") %>% 
  left_join(race_2010, by = "name")
# map
mapview(lowest_EISF_tract_2013to2017)
# join w/ race data
data1 <- lowest_EISF_tract_2013to2017[ , 4:10] %>% st_drop_geometry()
# plot
data <- data.frame(
  name = c(colnames(data1)) ,  
  value = c(as.numeric(data1[1, ]))
  )
# ..
ggplot(data, aes(x = name, y = value, fill = name)) + 
  geom_bar(stat = "identity") +
  ggtitle("Tract with LOWEST EISF") +
  xlab("Race Category") +
  ylab("Percentage of Population")


# LEST efficient tract:
highest_EISF_tract_2013to2017 <- beim_2013to2017_avgs %>% 
  filter(name == "06013314104") %>% 
  left_join(race_2010, by = "name")
# map
mapview(highest_EISF_tract_2013to2017)
# join w/ race data
data1 <- highest_EISF_tract_2013to2017[ , 4:10] %>% st_drop_geometry()
# plot
data <- data.frame(
  name = c(colnames(data1)) ,  
  value = c(as.numeric(data1[1, ]))
  )
# ..
ggplot(data, aes(x = name, y = value, fill = name)) + 
  geom_bar(stat = "identity") +
  ggtitle("Tract with HIGHEST EISF") +
  xlab("Race Category") +
  ylab("Percentage of Population")

# TODO: could also find Q1 and Q3 (circle back later)
```

  As we can see above..


---

Scrap Code Below:












