---
title: "HendricksonSwen_Project_DATA-IN"
author: "Swen Hendrickson"
date: "12/12/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r 1. setup, include = F}
knitr::opts_chunk$set(echo = F, warning = F, message = F)

# Clear all variables, close all plots, and clear the console, load libraries
rm(list=ls())
graphics.off()
cat("\014")

setwd("~/GitHub/swehen17.github.io/218X/Project")

library(censusapi)
library(tidyverse)
library(tigris)
library(sf)
library(leaflet)
library(mapview)
library(plotly)
library(data.table)
library(viridis)
library(hrbrthemes)
library(janitor)

Sys.setenv(CENSUS_KEY="c8aa67e4086b4b5ce3a8717f59faa9a28f611dab")
```

Pasted in from v2:
(methodology below)
```{r EISF}
residential_energy_intensity <- read_csv("usage_bld_btu.csv") %>%
  filter(
    usetype == "single_family",
  ) %>%
  select(
    geo_id,
    usetype,
    year,
    sqft,
    usage,
    name
  )

# only include census tracts, remove other geographical aggregations (cities, zips, etc.)
residential_energy_intensity <- residential_energy_intensity[1:7920,]

# sum(is.na(residential_energy_intensity$usage))
# 459 NA values for usage

# remove 'masked' values and create energy intensity calc
residential_energy_intensity <- residential_energy_intensity %>%
  filter(
    usage != "masked"
  ) %>%
  mutate(
    energy_intensity = as.numeric(usage) / sqft
  )

# map energy intensity
#

ca_tracts <- readRDS("ca_tracts.rds") %>%
  st_transform(crs= 4326)

# next steps: follow steps in A5

res_eng_int_2017 <- residential_energy_intensity %>%
  filter(year == 2017)

# get just tracts in my energy dataset
bay_tracts_need <- unique(residential_energy_intensity$name)

saveRDS(bay_tracts_need, file = "bay_tracts_need.rds")

bay_tracts_use_geo <- ca_tracts %>%
  filter(GEOID %in% bay_tracts_need) %>%
  select(GEOID, geometry) %>%
  transmute(name = GEOID) %>%
  st_transform(crs = 4326)

saveRDS(bay_tracts_use_geo, "bay_tracts_use_geo.rds")

bay_eng_int_map <- residential_energy_intensity %>%
  left_join(bay_tracts_use_geo, by = "name") %>%
  select(year, energy_intensity, geometry, name) %>%
  st_as_sf() %>%
  filter(energy_intensity != "Inf") %>%
  st_transform(crs = 4326)

saveRDS(bay_eng_int_map, "bay_eng_int_map.rds")

beim_2013 <- bay_eng_int_map %>%
  filter(year == 2013) %>%
  st_transform(crs = 4326)

saveRDS(beim_2013, "beim_2013.rds")

# avg of 2013-2017
beim_2013to2017_avgs <- as.data.frame(bay_eng_int_map) %>%
  group_by(
    name
  ) %>%
  summarise(
    avg = sum(energy_intensity / 5)
  ) %>% left_join(
    bay_tracts_use_geo, by = "name"
  ) %>%
  st_as_sf()

saveRDS(beim_2013to2017_avgs, file = "beim_2013to2017_avgs.rds")
```
Methodology of above chunk:
	- Took in data (from Energy Atlas) which included utility energy consumption 
	data for buildings of all types, aggregated by census tract
	- Filtered for residential single-family homes (data included multi family, 
	commercial, industrial, etc.)
	- Divided total usage by built SF to get energy intensity
	- mapped an individual year, then
	- Took average of year 2013-2017
  - Mapped by energy intensity (Btu/SF), darker blue being higher

```{r ECPC}
# now get energy consumption per capita (from the same data set [Energy Atlas])
ECPC <- read_csv("usage_bld_btu.csv") %>%
  select(usetype, year, sqft, usage_percap, name) %>%
  filter(
    usetype %in% c('res_total', 'single_family'),
    name %in% unique(residential_energy_intensity$name)) %>%
  mutate(
    singfam_perc = sqft / shift(sqft, 1L, type = "lag"),
    usage_percap = singfam_perc * shift(usage_percap, 1L, type = "lag")
    ) %>%
  filter(usetype == "single_family") %>%
  select(year, usage_percap, name)

saveRDS(ECPC, "ECPC.rds")

ECPC_2013 <- ECPC %>% 
  filter(year == 2013, usage_percap!= "masked") %>% 
  transmute(
    ecpc = usage_percap,
    name = name
  ) %>% 
  left_join(bay_tracts_use_geo, by = "name") %>% 
  st_as_sf()

saveRDS(ECPC_2013, "ECPC_2013.rds")
```
Methodology of above:
- only had population for total residential (includes mutli family, etc)
- but had SF data for single family homes. got percentage that single family
homes make up of total residential, to multiply by the usage per capita for
total residential to get usage per capita for single family homes
- helps to control for neighborhoods that might have different proportions 
of homes vs. multifamily complexes, mobile homes, etc.

```{r ECPH}
# energy consumption per household
ECPH <- read_csv("usage_bld_btu.csv") %>%
  select(usetype, year, usage, name) %>%
  filter(
    usetype %in% c('res_total'),
    name %in% unique(residential_energy_intensity$name))

ECPH_2013 <- ECPH %>% 
  filter(year == 2013, usage != "masked") %>% 
  left_join(units_in_structure_2013, by = "name") %>% 
  select( -c("usetype", "year")) %>% 
  transmute(
    ecph = (as.numeric(usage) * perc_1_unit) / (total_1_units),
    name = name
  ) %>% 
  left_join(bay_tracts_use_geo, by = "name") %>% 
  st_as_sf()
    
saveRDS(ECPH_2013, "ECPH_2013.rds")
```

```{r sociodem: INCOME}
# ACS 5-year estimates for years 2013 - 2017
income_2013 <-
  read_csv("data/income/ACSDT5Y2013.B19001_data_with_overlays_2021-12-04T162248.csv") %>%
  mutate(name = substr(GEO_ID, 10, 20)) %>%
  select(-c(ends_with("M"), "NAME", "GEO_ID")) %>%
  filter(name %in% unique(residential_energy_intensity$name)) %>%
  transmute(
    'total' = as.numeric(B19001_001E),
    '5k' = as.numeric(B19001_002E),
    '12.5k' = as.numeric(B19001_003E),
    '17.5k' = as.numeric(B19001_004E),
    '22.5k' = as.numeric(B19001_005E),
    '27.5k' = as.numeric(B19001_006E),
    '32.5k' = as.numeric(B19001_007E),
    '37.5k' = as.numeric(B19001_008E),
    '42.5k' = as.numeric(B19001_009E),
    '47.5k' = as.numeric(B19001_010E),
    '55k' = as.numeric(B19001_011E), # PATTERN STARTS CHANGING
    '67.5k' = as.numeric(B19001_012E),
    '87.5k' = as.numeric(B19001_013E),
    '112.5k' = as.numeric(B19001_014E),
    '137.5k' = as.numeric(B19001_015E),
    '175k' = as.numeric(B19001_016E),
    '250k' = as.numeric(B19001_017E), # $200k 'or 'more'. is there a better est for this?
    name = name) %>%
  mutate(
    weighted_avg_income =
      1000 * (`5k` * 5 +
    `12.5k` * 12.5 +
    `17.5k` * 17.5 +
    `22.5k` * 22.5 +
    `27.5k` * 27.5 +
    `32.5k` * 32.5 +
    `37.5k` * 37.5 +
    `42.5k` * 42.5 +
    `47.5k` * 47.5 +
    `55k` * 55 +
    `67.5k` * 67.5 +
    `87.5k` * 87.5 +
    `112.5k` * 112.5 +
    `137.5k` * 137.5 +
    `175k` * 175 +
    `250k` * 250)
    / total)

saveRDS(income_2013, "income_2013.rds")

##

income_2014 <-
  read_csv("data/income/ACSDT5Y2014.B19001_data_with_overlays_2021-12-04T162248.csv") %>%
  mutate(name = substr(GEO_ID, 10, 20)) %>%
  select(-c(ends_with("M"), "NAME", "GEO_ID", "B19001_001E")) %>%
  filter(name %in% unique(residential_energy_intensity$name))

income_2015 <-
  read_csv("data/income/ACSDT5Y2015.B19001_data_with_overlays_2021-12-04T162248.csv") %>%
  mutate(name = substr(GEO_ID, 10, 20)) %>%
  select(-c(ends_with("M"), "NAME", "GEO_ID", "B19001_001E")) %>%
  filter(name %in% unique(residential_energy_intensity$name))

income_2016 <-
  read_csv("data/income/ACSDT5Y2016.B19001_data_with_overlays_2021-12-04T162248.csv") %>%
  mutate(name = substr(GEO_ID, 10, 20)) %>%
  select(-c(ends_with("M"), "NAME", "GEO_ID", "B19001_001E")) %>%
  filter(name %in% unique(residential_energy_intensity$name))

income_2017 <-
  read_csv("data/income/ACSDT5Y2017.B19001_data_with_overlays_2021-12-04T162248.csv") %>%
  mutate(name = substr(GEO_ID, 10, 20)) %>%
  select(-c(ends_with("M"), "NAME", "GEO_ID", "B19001_001E")) %>%
  filter(name %in% unique(residential_energy_intensity$name))
```

```{r SELECTED HOUSING CHARACTERISTICS DATA}
sel_house_chars_2013 <- read_csv(
  "data/selected_housing_characteristics/ACSDP5Y2013.DP04_data_with_overlays_2021-12-12T194948.csv")

    # 61, 65, 69, 73, 77, 81, 85, 89, 93, 97,   # year structure built
    # 21, 25, 29,   # units in structure
    # 173, 177, 181, 185, 189,   # housing tenure
    # 193, 197, 201, 205, 209, 213, 217,   # year householder moved in
    # 313, 317, 321, 325, 329, 333, 337, 341, 345, 349,   # Value
    # 565   # GEO_ID

##

sel_house_chars_2014 <- read_csv(
  "data/selected_housing_characteristics/ACSDP5Y2014.DP04_data_with_overlays_2021-12-12T194948.csv")

sel_house_chars_2015 <- read_csv(
  "data/selected_housing_characteristics/ACSDP5Y2015.DP04_data_with_overlays_2021-12-12T194948.csv")

sel_house_chars_2016 <- read_csv(
  "data/selected_housing_characteristics/ACSDP5Y2016.DP04_data_with_overlays_2021-12-12T194948.csv")

sel_house_chars_2017 <- read_csv(
  "data/selected_housing_characteristics/ACSDP5Y2017.DP04_data_with_overlays_2021-12-12T194948.csv")

# bay_tracts_need <- readRDS("")

# build and filter 2013, then make loop to put all years into one
# can then filter more efficiently (fewer lines of code)

# years <- 2013:2017
# 
# sel_house_chars <- NULL
# 
# for(year in years) {
#   
#       filename <-
#         paste0(
#           "data/selected_housing_characteristics/ACSDP5Y",
#           year,
#           ".DP04_data_with_overlays_2021-12-12T194948.csv"
#         )
# 
#       print(filename)
#       
#       temp <- read_csv(filename)
# 
#       sel_house_chars <-
#         rbind(sel_house_chars, temp)
# }
```

```{r sociodem: TENURE}
# Tenure (Own vs. Rent)
housing_tenure_2013 <- sel_house_chars_2013 %>% 
  select(173, 177, 181, 185, 189, 565) %>% 
  row_to_names(row_number = 1) %>% 
  transmute(
    total_units = as.numeric(
      `Estimate!!HOUSING TENURE!!Occupied housing units`),
    owned_units = as.numeric(
      `Estimate!!HOUSING TENURE!!Occupied housing units!!Owner-occupied`),
    rented_units = as.numeric(
      `Estimate!!HOUSING TENURE!!Occupied housing units!!Renter-occupied`),
    avg_occ_owned_unit = as.numeric(
      `Estimate!!HOUSING TENURE!!Average household size of owner-occupied unit`),
    avg_occ_rented_unit = as.numeric(
      `Estimate!!HOUSING TENURE!!Average household size of renter-occupied unit`),
    name = substr(id, 10, 20)  
  ) %>% 
  transmute(
    "owned_units" = owned_units,
    "rented_units" = rented_units,
    "own_occ" = owned_units * avg_occ_owned_unit,
    "rent_occ" = rented_units * avg_occ_rented_unit,
    name = name
  ) %>% 
  transmute(
    owned_units_perc = owned_units / (owned_units + rented_units),
    owned_occ_perc = own_occ / (own_occ + rent_occ),
    name = name
  ) %>% 
  filter(
    name %in% unique(residential_energy_intensity$name)
    )

saveRDS(housing_tenure_2013, file = "housing_tenure_2013.rds")
```

```{r sociodem: YEAR MOVED IN}
# Number of Years Living in Home
years_in_home_2013 <- sel_house_chars_2013 %>% 
  select(193, 197, 201, 205, 209, 213, 217, 565) %>% 
  row_to_names(row_number = 1) %>% 
  transmute(
    "total" = as.numeric(`Estimate!!YEAR HOUSEHOLDER MOVED INTO UNIT!!Occupied housing units`),
    "5yrs" = as.numeric(`Estimate!!YEAR HOUSEHOLDER MOVED INTO UNIT!!Occupied housing units!!Moved in 2010 or later`),
    "15yrs" = as.numeric(`Estimate!!YEAR HOUSEHOLDER MOVED INTO UNIT!!Occupied housing units!!Moved in 2000 to 2009`),
    "25yrs" = as.numeric(`Estimate!!YEAR HOUSEHOLDER MOVED INTO UNIT!!Occupied housing units!!Moved in 1990 to 1999`),
    "35yrs" = as.numeric(`Estimate!!YEAR HOUSEHOLDER MOVED INTO UNIT!!Occupied housing units!!Moved in 1980 to 1989`),
    "45yrs" = as.numeric(`Estimate!!YEAR HOUSEHOLDER MOVED INTO UNIT!!Occupied housing units!!Moved in 1970 to 1979`),
    "55yrs" = as.numeric(`Estimate!!YEAR HOUSEHOLDER MOVED INTO UNIT!!Occupied housing units!!Moved in 1969 or earlier`),
    "name" = substr(id, 10, 20)
  ) %>% 
  filter(name %in% unique(residential_energy_intensity$name)) %>% 
  mutate(
    weighted_avg_years_in_home = (
          `5yrs` * 5 +
          `15yrs` * 15 +
          `25yrs` * 25 +
          `35yrs` * 35 +
          `45yrs` * 45 +
          `55yrs` * 55)
          / total
  ) %>% 
  select(name, weighted_avg_years_in_home)

saveRDS(years_in_home_2013, file = "years_in_home_2013.rds")
```

```{r sociodem: RACE}
race_2010 <-
  read_csv(
    "data/race/DECENNIALSF12010.P8_data_with_overlays_2021-12-05T025058.csv")
race_2010 <- race_2010[2:8058 , 1:10] %>%
  mutate(name = substr(GEO_ID, 10, 20)) %>% select(-c(GEO_ID, P008002)) %>%
  transmute(
    'Total' = as.numeric(P008001),
    'White' = as.numeric(P008003),
    'Black' = as.numeric(P008004),
    'Native_American' = as.numeric(P008005),
    'Asian' = as.numeric(P008006),
    'Pacific_Islander' = as.numeric(P008007),
    'Other' = as.numeric(P008008),
    'Multi-Racial' = as.numeric(P008009),
    'name' = name
  ) %>%
  filter(name %in% unique(residential_energy_intensity$name))

race_2010 <- race_2010 %>%
  transmute(
    'White' = White / Total,
    'Black' = Black / Total,
    'Native_American' = Native_American / Total,
    'Asian' = Asian / Total,
    'Pacific_Islander' = Pacific_Islander / Total,
    'Other' = Other / Total,
    'Multi-Racial' = `Multi-Racial` / Total,
    'name' = name
  ) %>% filter(!is.na(.))
# this gives us census tracts with race breakdowns by percentage

saveRDS(race_2010, file = "race_2010.rds")
```

```{r UNITS IN STRUCTURE}
units_in_structure_2013 <- sel_house_chars_2013 %>% 
  select(21, 25, 29, 565) %>% 
  row_to_names(row_number = 1) %>% 
  transmute(
    "Total housing units" = as.numeric(`Estimate!!UNITS IN STRUCTURE!!Total housing units`),
    "1-unit, detached" = as.numeric(`Estimate!!UNITS IN STRUCTURE!!Total housing units!!1-unit, detached`),
    "1-unit, attached" = as.numeric(`Estimate!!UNITS IN STRUCTURE!!Total housing units!!1-unit, attached`),
    name = substr(id, 10, 20)               
  ) %>% 
  transmute(
    "total_1_units" = (`1-unit, detached` + `1-unit, attached`),
    "perc_1_unit" = (`1-unit, detached` + `1-unit, attached`) / `Total housing units`,
    name = name
  )
```

```{r AVG HOUSE SIZE}
avg_house_size_2013 <- units_in_structure_2013 %>% 
  left_join(residential_energy_intensity, by = "name") %>% 
  filter(
    name %in% unique(residential_energy_intensity$name),
    year == 2013) %>% 
  select(sqft, total_1_units, name) %>% 
  transmute(
    avg_house_size = sqft / total_1_units,
    name = name
  ) 
  # %>% 
  # transmute(
  #   avg_house_size = 
  #     ifelse(
  #       avg_house_size < 500 & avg_house_size > 10000,
  #       "NA",
  #       avg_house_size
  #         ),
  #   name = name
  # )

saveRDS(avg_house_size_2013, file = "avg_house_size_2013.rds")
```


```{r VALUE}
VALUE_2013 <- sel_house_chars_2013 %>% 
  select(313, 317, 321, 325, 329, 333, 337, 341, 345, 565) %>% 
  row_to_names(row_number = 1) %>% 
  transmute(
    'total' = as.numeric(`Estimate!!VALUE!!Owner-occupied units`),
    '25k' = as.numeric(`Estimate!!VALUE!!Owner-occupied units!!Less than $50,000`),
    '75k' = as.numeric(`Estimate!!VALUE!!Owner-occupied units!!$50,000 to $99,999`),
    '125k' = as.numeric(`Estimate!!VALUE!!Owner-occupied units!!$100,000 to $149,999`),
    '175k' = as.numeric(`Estimate!!VALUE!!Owner-occupied units!!$150,000 to $199,999`),
    '250k' = as.numeric(`Estimate!!VALUE!!Owner-occupied units!!$200,000 to $299,999`),
    '400k' = as.numeric(`Estimate!!VALUE!!Owner-occupied units!!$300,000 to $499,999`),
    '750k' = as.numeric(`Estimate!!VALUE!!Owner-occupied units!!$500,000 to $999,999`),
    '1250k' = as.numeric(`Estimate!!VALUE!!Owner-occupied units!!$1,000,000 or more`),
    name = substr(id, 10, 20)
  ) %>% 
  filter(name %in% unique(residential_energy_intensity$name)) %>% 
  mutate(
    weighted_avg_value =
      1000 * 
      (
      `25k` * 25 +
      `75k` * 75 +
      `125k` * 125 +
      `175k` * 175 +
      `250k` * 250 +
      `400k` * 400 +
      `750k` * 750 +
      `1250k` * 1250
      ) / total
  ) %>% 
  select(name, weighted_avg_value)

# note these are all owner-occupied units

saveRDS(VALUE_2013, file = "VALUE_2013.rds")
```




































